---
title: "Practical Machine Learning - Barbell Lift"
author: "Cassie Breen"
date: "May 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

For this project, we are given data from accelerometers on the belt, forearm, arm, and dumbell of 6 research study participants. Our training data consists of accelerometer data and a label identifying the quality of the activity the participant was doing. Our testing data consists of accelerometer data without the identifying label. Our goal is to predict the labels for the test set observations.

Below is the code I used when creating the model, estimating the out-of-sample error, and making predictions. I also include a description of each step of the process.

## Data Preparation

```{r, cache = T}
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(RColorBrewer)
library(randomForest)
library(knitr)
```

Getting and loading the data:

```{r, cache = T}
set.seed(12345)

trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

Partioning the training set into two

```{r, cache = T}
inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
dataTraining <- training[inTrain, ]
dataTesting <- training[-inTrain, ]
dim(dataTraining); dim(dataTesting)
```

## Cleaning the data
Remove NearZeroVariance variables

```{r, cache = T}
nzv <- nearZeroVar(dataTraining, saveMetrics=TRUE)
dataTraining <- dataTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(dataTesting,saveMetrics=TRUE)
dataTesting <- dataTesting[,nzv$nzv==FALSE]
```

Remove the first column of the dataTraining data set

```{r, cache = T}
dataTraining <- dataTraining[c(-1)]
```

Clean variables with more than 60% NA

```{r, cache = T}
trainingV3 <- dataTraining
for(i in 1:length(dataTraining)) {
    if( sum( is.na( dataTraining[, i] ) ) /nrow(dataTraining) >= .7) {
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(dataTraining[i]), names(trainingV3)[j]) ) == 1)  {
                trainingV3 <- trainingV3[ , -j]
            }   
        } 
    }
}
```

# Set back to the original variable name

```{r, cache = T}
dataTraining <- trainingV3
rm(trainingV3)
```

Transform the dataTesting and testing data sets

```{r, cache = T}
clean1 <- colnames(dataTraining)
clean2 <- colnames(dataTraining[, -58])  # remove the classe column
dataTesting <- dataTesting[clean1]         # allow only variables in myTesting that are also in myTraining
testing <- testing[clean2]             # allow only variables in testing that are also in myTraining

dim(dataTesting)
dim(testing)
```

Coerce the data into the same type

```{r, cache = T}
for (i in 1:length(testing) ) {
    for(j in 1:length(dataTraining)) {
        if( length( grep(names(dataTraining[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(dataTraining[i])
        }      
    }      
}

# To get the same class between testing and dataTraining
testing <- rbind(dataTraining[2, -58] , testing)
testing <- testing[-1,]
```

##Prediction with Decision Trees

```{r, cache = T}
set.seed(12345)
modFitA1 <- rpart(classe ~ ., data=dataTraining, method="class")
fancyRpartPlot(modFitA1)
```

```{r, cache = T}
predictionsA1 <- predict(modFitA1, dataTesting, type = "class")
cmtree <- confusionMatrix(predictionsA1, dataTesting$classe)
cmtree
```

```{r, cache = T}
plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```

##Prediction with Random Forests
```{r, cache = T}
set.seed(12345)
modFitB1 <- randomForest(classe ~ ., data=dataTraining)
predictionB1 <- predict(modFitB1, dataTesting, type = "class")
cmrf <- confusionMatrix(predictionB1, dataTesting$classe)
cmrf
plot(modFitB1)
```
```{r, cache = T}
plot(cmrf$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```

##Prediction with Generalized Boosted Regression
```{r, cache = T}
set.seed(12345)
fitControl <- trainControl(method = "repeatedcv",
                           number = 5,
                           repeats = 1)

gbmFit1 <- train(classe ~ ., data=dataTraining, method = "gbm",
                 trControl = fitControl,
                 verbose = FALSE)


gbmFinMod1 <- gbmFit1$finalModel

gbmPredTest <- predict(gbmFit1, newdata=dataTesting)
gbmAccuracyTest <- confusionMatrix(gbmPredTest, dataTesting$classe)
gbmAccuracyTest
plot(gbmFit1, ylim=c(0.9, 1))
```

##Predicting Results on the Test Data
Random Forests gave an Accuracy in the myTesting dataset of 99.89%, which was more accurate that what I got from the Decision Trees or GBM. The expected out-of-sample error is 100-99.89 = 0.11%.
```{r, cache = T}
predictionB2 <- predict(modFitB1, testing, type = "class")
predictionB2
```

```{r, cache = T}
# Write the results to a text file for submission
pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}# pml_write_files(predictionB2)
```